---
title: "Sales Performance and Profit Optimization Analysis"
author: "Ajik Shadrach"
date: "2026-03-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Sales Performance and Profit Optimization Analysis

This report analyzes retail sales data to evaluate business performance and identify opportunities to improve profitability.

---

# Load Required Libraries

```{r libraries}
library(tidyverse)
library(lubridate)
library(scales)
library(janitor)
```

---

# Import Dataset

```{r import-data}
sales_data <- read.csv("superstore_sales.csv", stringsAsFactors = FALSE)

sales_data <- clean_names(sales_data)

str(sales_data)
```

---

# Data Cleaning

```{r data-cleaning}
sales_data$order_date <- mdy(sales_data$order_date)
sales_data$ship_date <- mdy(sales_data$ship_date)

colSums(is.na(sales_data))

sales_data <- distinct(sales_data)

sales_data$sales <- as.numeric(sales_data$sales)
sales_data$profit <- as.numeric(sales_data$profit)
sales_data$discount <- as.numeric(sales_data$discount)
sales_data$quantity <- as.numeric(sales_data$quantity)
```

---

# Key Performance Indicators (KPIs)

```{r kpi-calculation}
total_sales <- sum(sales_data$sales)
total_profit <- sum(sales_data$profit)
total_orders <- n_distinct(sales_data$order_id)

profit_margin <- total_profit / total_sales

kpi_table <- data.frame(
Total_Sales = total_sales,
Total_Profit = total_profit,
Total_Orders = total_orders,
Profit_Margin = profit_margin
)

kpi_table
```

---

# Regional Sales Performance

```{r region-summary}
region_summary <- sales_data %>%
group_by(region) %>%
summarise(
total_sales = sum(sales),
total_profit = sum(profit)
)

region_summary
```

## Sales by Region

```{r region-plot}
ggplot(region_summary, aes(x = reorder(region, total_sales), y = total_sales)) +
geom_col() +
coord_flip() +
scale_y_continuous(labels = comma) +
labs(
title = "Total Sales by Region",
x = "Region",
y = "Total Sales"
)
```

---

# Profit Margin by Category

```{r category-summary}
category_profit <- sales_data %>%
group_by(category) %>%
summarise(
total_sales = sum(sales),
total_profit = sum(profit)
) %>%
mutate(profit_margin = total_profit / total_sales)

category_profit
```

## Category Profit Margin Plot

```{r category-plot}
ggplot(category_profit, aes(x = category, y = profit_margin)) +
geom_col() +
scale_y_continuous(labels = percent) +
labs(
title = "Profit Margin by Category",
x = "Category",
y = "Profit Margin"
)
```

---

# Customer Segmentation Analysis

```{r segment-summary}
segment_summary <- sales_data %>%
group_by(segment) %>%
summarise(
total_sales = sum(sales),
total_profit = sum(profit),
total_orders = n()
)

segment_summary
```

## Sales by Customer Segment

```{r segment-plot}
ggplot(segment_summary, aes(x = segment, y = total_sales)) +
geom_col() +
scale_y_continuous(labels = comma) +
labs(
title = "Sales by Customer Segment",
x = "Customer Segment",
y = "Total Sales"
)
```

---

# Discount vs Profit Relationship

```{r discount-profit}
ggplot(sales_data, aes(x = discount, y = profit)) +
geom_point(alpha = 0.4) +
geom_smooth(method = "lm") +
labs(
title = "Discount vs Profit Relationship",
x = "Discount",
y = "Profit"
)
```

---

# Key Business Insights

- Sales performance varies significantly across regions.
- Certain product categories generate higher profit margins than others.
- Customer segments contribute differently to total sales.
- Higher discounts tend to reduce profitability.

---

# Conclusion

The analysis highlights opportunities for improving profitability by optimizing discount strategies and focusing on high-performing product categories and customer segments.
